Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-05-03 papers/halfdisk/halfdisk.tex
Further edits

The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{P_n^{(b,b)}\}$ are not orthonormal, but have (squared) norms given by the integral of their respective weight functions we denote as
\begin{align*}
	\normgenjac^{(a,b)} := \int_0^1 \: \genjacw^{(a,b)}(x) \: \D x, \quad \normjac^{(b)} := \int_{-1}^1 \: w_P^{(b)}(y) \: \D y,
\end{align*}
i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}}^2 = \normgenjac^{(a,b)}$ $\forall n = 0,1,2,\dots$ and $\norm{P_{n}^{(b,b)}}_{w_P^{(b)}}^2 = \normjac^{(b)}$ $\forall n = 0,1,2,\dots$.

For clarity, we use the notation that 
\begin{align*}
	&\ip<p, q>_{\genjacw^{(a,b)}} := \int_0^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x, \quad
		\ip<p, q>_{w_P^{(b)}} := \int_{-1}^1 p(y) \: q(y) \: w_P^{(b)}(y)\: \D y, \\
	&\norm{p}_{\genjacw^{(a,b)}}^2 := \ip<p, p>_{\genjacw^{(a,b)}}, \quad
		\norm{p}_{w_P^{(b)}}^2 := \ip<p, p>_{w_P^{(b)}}.
\end{align*}

@@

\begin{lemma}
$\hdopnk^{(a,b)}(x,y)$ satisfy the following 3-term recurrences:
\begin{align*}
x \hdopnk^{(a,b)}(x,y) &= \alphaab_{n,k,1} \: \hdop_{n-1, k}^{(a,b)}(x, y) + \alphaab_{n,k,2} \: \hdop_{n, k}^{(a,b)}(x, y) + \alphaab_{n+1,k,1} \: \hdop_{n+1, k}^{(a,b)}(x, y), \\
y \hdopnk^{(a,b)}(x,y) &= \betaab_{n,k,1} \: \hdop_{n-1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,2} \: \hdop_{n-1, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,3} \: \hdop_{n, k-1}^{(a,b)}(x, y) + \betaab_{n,k,4} \: \hdop_{n, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,5} \: \hdop_{n+1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,6} \: \hdop_{n+1, k+1}^{(a,b)}(x, y),
\end{align*}
for \((x,y) \in \Omega\), where
\begin{align*}
\alphaab_{n,k,1} &:= \beta_{n-k-1}^{(a, b+k+\half)}, \qquad \alphaab_{n,k,2} := \alpha_{n-k}^{(a, b+k+\half)} \\
\betaab_{n,k,1} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,2} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-2}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
\betaab_{n,k,3} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+1}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,4} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-1}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
\betaab_{n,k,5} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+2}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,6} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}}. 
\end{align*}

@@

\begin{proof}
The 3-term recurrence for multiplication by $x$ follows from equation (\ref{eqn:Hrecurrence}). For the recurrence for multiplication by $y$, first note that, using a change of variable $t = \frac{y}{\rho(x)}$,
\begin{align*}
	\norm{\hdopnkab}_\Wab^2 &=  \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \rho(x)^{2k} \: {P_k^{(b,b)}\fpr(\frac{y}{\rho(x)})^2} \: \Wab(x,y) \: \D y \: \D x \nonumber \\
	&= \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: {P_k^{(b,b)}\fpr(\frac{y}{\rho(x)})^2} \: \genjacw^{(a,b+k)}(x) \: w_P^{(b)}\fpr(\frac{y}{\rho(x)}) \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \genjacw^{(a,b+k+\half)}(x) \: \D x \Big) \: \Big( \int_\gamma^\delta \: {P^{(b,b)}}(t)^2 \: w_P^{(b)}(t) \: \D t \Big) \nonumber \\
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)}.
\end{align*}
where we use the notation that $\ip<p, q>_{\Wab} := \iint_\Omega p(x,y) \: q(x,y) \: \Wab(x,y) \: \D y \: \D x$ and $\norm{p}_{W^{(a,b)}}^2 := \ip<p, p>_{\Wab}$ .

Finally, note that \(\ip<y \hdopnkab, \hdopmjab>_\Wab = 0\) for \(m < n-1\) (and for \(m > n+2\)). Thus for \(m = n-1, n, n+1,\) \(j = 0,\dots,m:\) using the same change of variable:
\begin{align*}
	\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \rho(x)^{k+j+2} \: \genjacw^{(a, b)}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: w_P^{(b)}(t) \: \D t \Big) \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \genjacw^{(a, b+1+\half(k+j))}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: w_P^{(b)}(t) \: \D t \Big) \\
	&= 
	\begin{cases}
    		\delta_k^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
		\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k+\half)}} \quad& \text{if } j = k-1 \\
		0 & \text{otherwise}
      	\end{cases}.
\end{align*}

@@

Finally, note that the inner integral at $(\dagger)$ over $t$ would be zero if $y \mapsto f(x,y)$ for fixed $x$ was an odd function, due to the symmetry over the domain (i.e. since $\gamma = -\delta$ in the this case).
\end{proof}

@@

Fix \(a,b \in \R\). Then for any function \(f : \Omega \to \R\) of degree $N$ we can express \(f\) by
\begin{align*}
f(x,y) = \sum_{n=0}^N \bighdop_n^{(a,b)}(x,y)^\top \: \mathbf{f}_n
\end{align*}
where
\begin{align*}
\bighdop_n(x,y) &:= \begin{pmatrix}
		\hdop_{n,0}(x,y) \\
		\vdots \\
		\hdop_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N,
\end{align*}
and where
\begin{align*}
\mathbf{f}_n &:= \begin{pmatrix}
		f_{n,0} \\
		\vdots \\
		f_{n,n}
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N, \quad
f_{n,k} := \frac{\langle f, \: \hdopnk^{(a,b)} \rangle_{W^{(a,b)}}}{|| \hdopnk^{(a,b)} ||_{W^{(a,b)}}}
\end{align*}

Using the quadrature rule detailed in Section 4.2, we can calculate exactly the coefficients $f_{n,k}$: 
\begin{align*}
	\ip< f, \: \hdopnk^{(a,b)} >_{W^{(a,b)}} &= \half \: \sum_{j=1}^{N^2} w_j \: \big[ f(x_j, y_j) \: \hdopnk^{(a,b)}(x_j, y_j) +f(x_j, -y_j) \: \hdopnk^{(a,b)}(x_j, -y_j) \big], \\
	|| \hdopnk^{(a,b)} ||_{W^{(a,b)}}^2 &= \half \: \sum_{j=1}^{N^2} w_j \: \big[ \hdopnk^{(a,b)}(x_j, y_j)^2 + \hdopnk^{(a,b)}(x_j, -y_j)^2 \big].
\end{align*}

>>>>

The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{P_n^{(b,b)}\}$ are not orthonormal, but have (squared) norms given by the integral of their respective weight functions we denote as
\begin{align*}
	\normgenjac^{(a,b)} := \int_0^1 \: \genjacw^{(a,b)}(x) \: \D x, \quad \normjac^{(b)} := \int_{-1}^1 \: w_P^{(b)}(y) \: \D y,
\end{align*}
i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}}^2 = \normgenjac^{(a,b)}$ $\forall n = 0,1,2,\dots$ and $\norm{P_{n}^{(b,b)}}_{w_P^{(b)}}^2 = \normjac^{(b)}$ $\forall n = 0,1,2,\dots$. For clarity, we use the notation that 
\begin{align*}
	&\ip<p, q>_{\genjacw^{(a,b)}} := \int_0^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x, \quad
		\ip<p, q>_{w_P^{(b)}} := \int_{-1}^1 p(y) \: q(y) \: w_P^{(b)}(y)\: \D y, \\
	&\norm{p}_{\genjacw^{(a,b)}}^2 := \ip<p, p>_{\genjacw^{(a,b)}}, \quad
		\norm{p}_{w_P^{(b)}}^2 := \ip<p, p>_{w_P^{(b)}}.
\end{align*}
Further, we have that the norms of the 2D half-disk OPs for a particular parameter family $(a,b)$ are then given by
\begin{align}
	\norm{\hdopnkab}_\Wab^2 &=  \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \rho(x)^{2k} \: {P_k^{(b,b)}\fpr(\frac{y}{\rho(x)})^2} \: \Wab(x,y) \: \D y \: \D x \nonumber \\
	&= \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: {P_k^{(b,b)}\fpr(\frac{y}{\rho(x)})^2} \: \genjacw^{(a,b+k)}(x) \: w_P^{(b)}\fpr(\frac{y}{\rho(x)}) \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \genjacw^{(a,b+k+\half)}(x) \: \D x \Big) \: \Big( \int_\gamma^\delta \: {P^{(b,b)}}(t)^2 \: w_P^{(b)}(t) \: \D t \Big) \nonumber \\
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)} \label{eqn:normhdop}
\end{align}
where we have used the change of variable $t = \frac{y}{\rho(x)}$. Again, for clairty we use the notation that $\ip<p, q>_{\Wab} := \iint_\Omega p(x,y) \: q(x,y) \: \Wab(x,y) \: \D y \: \D x$ and $\norm{p}_{\Wab}^2 := \ip<p, p>_{\Wab}$ .

@@

\begin{lemma}
$\hdopnk^{(a,b)}(x,y)$ satisfy the following 3-term recurrences:
\begin{align*}
x \hdopnk^{(a,b)}(x,y) &= \alphaab_{n,k,1} \: \hdop_{n-1, k}^{(a,b)}(x, y) + \alphaab_{n,k,2} \: \hdop_{n, k}^{(a,b)}(x, y) + \alphaab_{n+1,k,1} \: \hdop_{n+1, k}^{(a,b)}(x, y), \\
y \hdopnk^{(a,b)}(x,y) &= \betaab_{n,k,1} \: \hdop_{n-1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,2} \: \hdop_{n-1, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,3} \: \hdop_{n, k-1}^{(a,b)}(x, y) + \betaab_{n,k,4} \: \hdop_{n, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,5} \: \hdop_{n+1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,6} \: \hdop_{n+1, k+1}^{(a,b)}(x, y),
\end{align*}
for \((x,y) \in \Omega\), where
\begin{align*}
	\alphaab_{n,k,1} &:= \beta_{n-k-1}^{(a, b+k+\half)} \\
	\alphaab_{n,k,2} &:= \alpha_{n-k}^{(a, b+k+\half)} \\
	\betaab_{n,k,1} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
	\betaab_{n,k,2} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-2}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
	\betaab_{n,k,3} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+1}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
	\betaab_{n,k,4} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-1}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
	\betaab_{n,k,5} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+2}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
	\betaab_{n,k,6} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}}. 
\end{align*}

@@

\begin{proof}
The 3-term recurrence for multiplication by $x$ follows from equation (\ref{eqn:Hrecurrence}). For the recurrence for multiplication by $y$, since $\{\hdopmjab\}$ for $m = 0,\dots,n+1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n+1$ polynomial, we can expand $y \: \hdopnkab(x,) = \sum_{m=0}^{n+1} \sum_{j=0}^m c_{m,j} \: \hdopmjab(x,y)$. These coefficients are given by
\begin{align*}
	c_{m,j} = \frac{\ip< y \: \hdopnkab, \hdopmjab >_{\Wab}}{\norm{\hdopmjab}^2_{\Wab}}.
\end{align*}
Recall from equation (\ref{eqn:normhdop}) that $\norm{\hdopnkab}_{\Wab}^2 = \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)}$. Then for $m = 0,\dots,n+1$, $j = 0,\dots,m$, using the change of variable $t = \frac{y}{\rho(x)}$:
\begin{align*}
	\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \rho(x)^{k+j+2} \: \genjacw^{(a, b)}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: w_P^{(b)}(t) \: \D t \Big) \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \genjacw^{(a, b+1+\half(k+j))}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: w_P^{(b)}(t) \: \D t \Big) \\
	&= 
	\begin{cases}
    		\delta_k^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
		\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k+\half)}} \quad& \text{if } j = k-1 \\
		0 & \text{otherwise}
      	\end{cases}.
\end{align*}
where $\ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}}$ and $\ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k+\half)}}$ are zero for $m < n-1$.

@@

Finally, note that the inner integral at $(\dagger)$ over $t$ would be zero if $y \mapsto f(x,y)$ for fixed $x$ was an odd function, due to the symmetry over the domain (i.e. since $\gamma = -\delta$).
\end{proof}

@@

Fix \(a,b \in \R\). Then for any function \(f : \Omega \to \R\) of degree $N$ we can express \(f\) by
\begin{align*}
f(x,y) = \sum_{n=0}^N \bighdop_n^{(a,b)}(x,y)^\top \: \mathbf{f}_n
\end{align*}
where
\begin{align*}
\bighdop_n(x,y) &:= \begin{pmatrix}
		\hdop_{n,0}(x,y) \\
		\vdots \\
		\hdop_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N,
\end{align*}
and where
\begin{align*}
\mathbf{f}_n &:= \begin{pmatrix}
		f_{n,0} \\
		\vdots \\
		f_{n,n}
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N, \quad
f_{n,k} := \frac{\langle f, \: \hdopnk^{(a,b)} \rangle_{W^{(a,b)}}}{\norm{\hdopnk^{(a,b)}}_{W^{(a,b)}}}
\end{align*}

Recall from (\ref{eqn:normhdop}) that $\norm{\hdopnkab}^2_\Wab = \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)}$. Using the quadrature rule detailed in Section 4.2 for the inner product, we can calculate exactly the coefficients $f_{n,k}$: 
\begin{align*}
	f_{n,k} &= \frac{1}{2 \: \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)}} \: \sum_{j=1}^{N^2} w_j \: \big[ f(x_j, y_j) \: \hdopnk^{(a,b)}(x_j, y_j) +f(x_j, -y_j) \: \hdopnk^{(a,b)}(x_j, -y_j) \big].
\end{align*}
